Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2012-10-02 Dail_Mad_ext.tex
transition probs

Recall that conventional state-space models assume that the
distribution for the initial time period,
$[N_1 | \bm{\theta}]$, is either the equilibrium
distribution, or has zero variance.
In contrast, \citet{dail_madsen:2011} proposed modeling $N_{i1}$
as either a Poisson or negative binomial random variable:
\begin{gather}
N_{i1} \sim \mathrm{Pois}(\Lambda) \nonumber \\
\text{or} \nonumber \\
N_{i1} \sim \mathrm{NB}(\Lambda, \alpha)
\label{eq:N1}
\end{gather}
%\begin{equation}
%  [N_{i1} | \Lambda] = \frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}.
%  \label{eq:N1br}
%\end{equation}
where $\Lambda_i$ is the expected abundance at site $i$ during year
1. In the Poisson case, the eqivalent expression of
Eq.~\ref{eq:ss2a} is $[N_{i1} | \Lambda] =
\frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}$.
%From hereout, we will reserve bracket notation for arbitrary
%probability distributions, and for simplicity, use the equivalent notation $N_{i1}
%\sim \text{Pois}(\Lambda_i)$ when referencing specific distributions.
The Poisson distribution assumes that the mean of $N_{i1}$ is equal to its
variance, whereas the negative binomial distribution, $N_{i1} \sim
\text{NB}(\Lambda, \alpha)$, allows the
variance to be greater than the mean with the amount of overdispersion
determined by the parameter $\alpha$.
Note that $N_{i1} \sim \text{Pois}(\Lambda)$ is equivalent to
$[N_{i1} | ] = \frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}$.

@@

In the absence
of movement, $\omega$ is exactly the probability of surviving from
year $t$ to $t-1$, and $\gamma$ is the per-capita birth rate. However,
in the more common scenario when immigration and emigration occur,
$\omega$ is the probability of surviving and not emigrating, and
$\gamma$ is the sum %\hl{[I think!]}
of the birth rate and the immigration rate,
i.e. the recruitment rate. In the former case where no movement
occurs, the DM model can directly estimate
demographich parameters, and this mechanistic description of
population change is much more appealing than classical
population growth models such as the logistic model. However, movement
complicates the interpretation of the the parameters, reducing the
connection to specific population processes. In such cases, the model
of dynamics shown in Eq.~\ref{eq:Nt} can be replaced with standard
population models as described in Sect~\ref{sec:ext}. Irrespective of
the model for population dynamics, it is worthwhile to note that the
discrete distributions used in the DM model avoid the issues
associated with the Gaussian distribution used to model process
variation in standard state-space models.

>>>>

Recall that conventional state-space models assume that the
distribution for the initial time period,
$[N_1 | \bm{\theta}]$, is either the equilibrium
distribution, or has zero variance.
In contrast, \citet{dail_madsen:2011} proposed modeling $N_{i1}$
as either a Poisson or negative binomial random variable:
\begin{gather}
N_{i1} \sim \mathrm{Pois}(\Lambda) \nonumber \\
\text{or} \nonumber \\
N_{i1} \sim \mathrm{NB}(\Lambda, \alpha)
\label{eq:N1}
\end{gather}
where $\Lambda_i$ is the expected abundance at site $i$ during year
1.
The Poisson distribution assumes that the mean of $N_{i1}$ is equal to its
variance, whereas the negative binomial distribution allows the
variance to be greater than the mean with the amount of overdispersion
determined by the parameter $\alpha$.
In the Poisson case, the eqivalent expression of
$[N_1 | \bm{\theta}]$ (Eq.~\ref{eq:ss2a}) is
%$[N_{i1} | \Lambda] = \frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}$.
\begin{equation}
  [N_{i1} | \Lambda] = \frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}.
  \label{eq:N1br}
\end{equation}

@@

To make the connection between conventional state-space models and the
DM model, we need to replace $[N_t|N_{t-1}, \bm{\Theta}]$,
Eq.~\ref{eq:ss2b}, with an expression derived from
Eq.~\ref{eq:Nt}. This requires summing over all possible values of $S$
and $G$, which is accomplished using the discrete convolution:
\begin{equation}
[N_{it}|N_{it-1}, \omega, \gamma] =
  \sum_{S_{it-1}=0}^{\min(N_{it},N_{it-1})}
  \binom{N_{it-1}}{S_{it-1}}\omega^{S_{it-1}}(1-\omega)^{N_{it-1}-S_{it-1}} \times
  \frac{\gamma^{N_{it}-S_{it-1}}e^{-\gamma}}{(N_{it}-S_{it-1})!}
  \label{eq:P}
\end{equation}
where here we assumed no dependenced of $\gamma$ on $N_{it-1}$.

In the absence
of movement, $\omega$ is exactly the probability of surviving from
year $t$ to $t-1$, and $\gamma$ is the per-capita birth rate (under
the geometric growth model). This ability to directly estimate
demographic parameters from simple count data is one of the DM model's
most appearling features. However,
in the more common scenario when immigration and emigration occur,
$\omega$ and $\gamma$ no longer can be intepretted as vital
rates. Rather, $\omega$ is the probability of surviving and not
emigrating, and
$\gamma$ is the sum of the birth rate and the immigration rate,
i.e. the recruitment rate. In such cases, we propose
replacing Eq.~\ref{eq:Nt} with standard
population models.

\begin{comment}
  Irrespective of the model for population dynamics, it is worthwhile
  to note that the discrete distributions used in the DM model avoid
  the issues associated with the Gaussian distribution used to model
  process variation in standard state-space models.
\end{comment}
